---
title: "Machine Learning Using Titanic Dataset (Exploratory Analysis)"
author: "Himansu Sahoo"
date: "October 15, 2015"
output: html_document
---

### Read the dataset
```{r}
#train_raw <- read.csv(file="../titanic-train.csv")
#test_raw <- read.csv(file="../titanic-test.csv")
train_raw <- read.csv(file="../titanic-train.csv", na.strings=c("", " ", "NA", "NAN"))
test_raw <- read.csv(file="../titanic-test.csv", na.strings=c("", " ", "NA", "NAN"))
# treat all empty values, NAN as NA's
```

#### Step 1 : check the dimensions, length, class of variables
```{r}
dim(train_raw)
# train_raw has 891 rows/observations and 12 variables.
#names(train_raw)
#sapply(train_raw, length) # all variables have same length
sapply(train_raw, class)
dim(test_raw)
# test_raw has 418 rows/observations and 11 variables.
#names(test_raw)
#sapply(test_raw, length) # all variables have same length
#sapply(test_raw, class)
#head(train_raw)
#head(test_raw)
```
Just looking at these, test\_raw has one variable less, Survived, which is the response/target variable. All variables have same length. test\_raw variables have same class as train\_raw.

#### Step 2 : `str()`, `summary()`, `head()`, `tail()` of the dataset
```{r}
#str(train_raw)
#str(test_raw)
#head(train_raw, n=100) # print first 100 rows
#summary(train_raw)
#summary(test_raw)
```

Looking at `str()` for the dataset four variables; PassengerId, Name, Ticket, Cabin are just unique identifiers for each person. They are not helpful for the model. remove the variables.
looking at `head(dataset, n=100)`, there are lots of data are missing in Cabin variable.
It will be helpful to read them as NA's by modifing the `read.csv()` command.

Looking at `summary(train_raw)`, we see Age has 177 NA's, Cabin has 687 NA's, Embarked has 2 NA's.

similarly in `summary(test_raw)`, we see Age has 86 NA's, Fare has 1 NA, Cabin has 327 NA's.

#### Step 3 : Data cleaning : Missing value treatment
```{r}
#colSums(is.na(train_raw)) # will shows NA's for each variable
#colSums(is.na(test_raw))
func_ratioNA <- function(x) length(x[x==""])/length(x)
perc_NA <- apply(train_raw, 2, FUN = func_ratioNA)
print(perc_NA) # percentage of NA's in the dataset
```

Age has 20\%, Cabin has 77\%, Embarked has 0.2\% missing values. Rules of thumb is to discard the variable having more than 30\% missing values.

```{r}
#names(train_raw)
train_cleaned <- train_raw[, -which(perc_NA>0.30)]
#names(train_cleaned) # Cabin is removed
cat("train_cleaned : dimension :  ", dim(train_cleaned) , "\n")
remove_trainvar <- names(train_cleaned) %in% c("PassengerId", "Name", "Ticket", "Cabin")
train_final <- train_cleaned[, !remove_trainvar]
#names(train_final) # "PassengerId", "Name", "Ticket", "Cabin" are removed
cat("train_final : dimension :  ", dim(train_final) , "\n")
apply(train_final, 2, FUN = func_ratioNA)
```

#### Step 4 : Data cleaning for test\_data
```{r}
remove_testvar <- names(test_raw) %in% c("PassengerId", "Name", "Ticket", "Cabin")
test_final <- test_raw[, !remove_testvar]
cat("test_final : dimension :  ", dim(test_final) , "\n")
apply(test_final, 2, FUN = func_ratioNA)
```

#### Step 5 : Data visualization

```{r echo=FALSE}
plotvar <- function(x){
            is_factor <- is.factor(x)
        if (is_factor == "FALSE"){
                par(mfrow=c(1, 2), pty = "s")
                hist(x, col="blue", breaks=50)
                hist_entries <- length(x)
                hist_mean <- mean(x, na.rm=TRUE)
                hist_med <- median(x, na.rm=TRUE)
                hist_sd <- sd(x, na.rm=TRUE)
                legend("topright", legend = c(paste("Entries = ", hist_entries), 
            paste("Mean =", round(hist_mean, 1)),
            paste("Median =",round(hist_med, 1)),
            paste("Std.Dev =", round(hist_sd, 1))))#  , bty = "n")

            boxplot(x, main="Box Plot", horizontal=TRUE, col="red")
        }else{
          par(mfrow=c(1, 1), pty = "s") # square plotting
            barplot(summary(x)) # summary will include NA's, while table don't
            legend("topright", legend = levels(x))          
        }
}

```

```{r}
plotvar(train_final$Pclass)
```
```{r fig.width=4, fig.height=4}
plotvar(train_final$Sex)
```
```{r}
plotvar(train_final$Age)
```
```{r}
plotvar(train_final$SibSp)
```
```{r}
plotvar(train_final$Parch)
```
```{r}
plotvar(train_final$Fare)
```
```{r fig.width=4, fig.height=4}
plotvar(train_final$Embarked)
```

#### Step 6 : Data cleaning : Checking Outliers
Looking at the box plot for Fare variable, there is lots of data outside the IQR.

```{r echo=FALSE}
perc_outlier <- function(x){
        if(class(x) == "factor"){
                ( length(x)/length(x) )
        }else{
                x <- x[!is.na(x)]
                not_outlier <- (x > (quantile(x, 0.25) - 1.5*IQR(x))) & (x< (quantile(x, 0.75) + 1.5*IQR(x)))
                length(x[!not_outlier])/length(x)
        }
}
```

```{r}
perc_outlier(train_final$Fare) # to check percentage of outlier
# 13\% of Fare data are outlier. Lets keep it now. we will study later.
```









