---
title: "Decision Tree using Titanic Dataset (CARET package)"
author: "Himansu Sahoo"
date: "December 29, 2015"
output: html_document
---

This is the example showing how to use decision tree algorithm in titanic dataset to predict survival rate of passengers.

link to kaggle competition : <https://www.kaggle.com/c/titanic>

### Step 1 : Getting the Data into R
```{r}
train_raw <- read.csv(file="../datasets/titanic-train.csv", na.strings=c(""," ", "NA", "NAN"))
test_raw <- read.csv(file="../datasets/titanic-test.csv", na.strings=c(""," ", "NA", "NAN"))
# stringsAsFactors=TRUE by default
# treat all empty values, NAN as NA's
```

```{r}
library(dplyr)
# convert into dplyr tbl format
train_raw <- tbl_df(train_raw)
test_raw <- tbl_df(test_raw)
```

#### Check the training dataset
```{r}
dim(train_raw)
# the dataset contains 891 rows/observations and 12 variables
#nrow(train_raw); ncol(train_raw)
#head(train_raw) # print first 6 rows
#tail(train_raw) # print last 6 rows
#str(train_raw) # will show class of each variable
# 'data.frame':    891 obs. of  12 variables:
#summary(train_raw)
glimpse(train_raw) # same as str() in base R
```

Looking at `str(train_raw)`  
Name       : Factor w/ 891 levels  
Ticket     : Factor w/ 681 levels  
Cabin      : Factor w/ 147 levels  

Note that Name and ticket are strings, they are read as factors.  
Looking at `summary(train_raw)`, Age has 177 NA's, Cabin has 687 NA's, Embarked has 2 NA's.

#### Check the testing dataset

```{r}
dim(test_raw)
#str(test_raw)
#'data.frame':    418 obs. of  11 variables:
#summary(test_raw)
glimpse(test_raw)
```

Looking at `str(test_raw)`  
Name       : Factor w/ 418 levels  
Ticket     : Factor w/ 363 levels  
Cabin      : Factor w/ 76 levels  

Looking at *summary(test_raw)*, Fare has 1 NA, Cabin has 327 NA's.  

#### Removing Variables
passengerId, Name are an unique identifier, remove this from the modeling
remove Ticket, Cabin
```{r}
train_final <- train_raw %>% select(-c(PassengerId, Name, Ticket, Cabin))
test_final <- test_raw
```

#### Missing Value treatment
```{r}
colSums(is.na(train_final))
train_final$Age[is.na(train_final$Age)] <- median(train_final$Age, na.rm=TRUE)

summary(train_final$Embarked)
#table() willn't print NA's
# maximum observations have S, replace NA's by S
train_final$Embarked[is.na(train_final$Embarked)] <- "S"
colSums(is.na(train_final))
```

Change the response variable to a factor variable, otherwise CART train function will give problem. It is thinking the problem as regression and outcome has only two levels.
Also, change the level names of the response variable, otherwise will give problem when calculating class probabilites.

```{r}
train_final$Survived <- as.factor(train_final$Survived)
#class(train_final$Survived) # factor
str(train_final$Survived) # Factor w/ 2 levels "0","1":
#levels(train_final$Survived)[levels(train_final$Survived)=="0"] <- "No"
#levels(train_final$Survived)[levels(train_final$Survived)=="1"] <- "Yes"
#levels(train_final$Survived)
```

#### Splitting data into training and validation set

We will use the `createDataPartition` function in `caret` package to split the `train_data` into 70\% train\_set and valid\_set. Set seed in the beginning to reproduce the same result in future.

```{r}
library(caret)
```

```{r}
set.seed(1099)
inTrain <- createDataPartition(y=train_final$Survived, p=0.70, list=FALSE)
train_set <- train_final[inTrain,]
valid_set <- train_final[-inTrain,]
```

```{r}
train_per <- (nrow(train_set)/nrow(train_final))*100
test_per <- (nrow(valid_set)/nrow(train_final))*100
cat("******** training dataset is : ", train_per, "% \n")
cat("******** testing dataset is : ", test_per, "% \n")
```

### Step 2 : Building a Decision Tree (CART) Model using CARET package

We will use the `caret` package in R to build the **decisionTree** model (CART) with method **rpart**.

```{r}
library(rpart)
dt_model <- train(Survived~., method="rpart", data=train_set)
```

After running the model, you will get the following Warnings. Make sure to change the target variable to a factor variable.

```
## Warning in train.default(x, y, weights = w, ...): You are trying to do
## regression and your outcome only has two possible values Are you trying to
## do classification? If so, use a 2 level factor as your outcome column.
```

```
## Warning in nominalTrainWorkflow(x = x, y = y, wts = weights, info =
## trainInfo, : There were missing values in resampled performance measures.
```

#### Print different Model functions

```{r}
class(dt_model) # "train"         "train.formula"
print(dt_model) # str(), summary() will print rubbish
names(dt_model)
ls(dt_model)
dt_model$method
dt_model$modelType
dt_model$call
dt_model$metric
dt_model$perfNames
dt_model$bestTune
dt_model$coefnames # will print long list
dt_model$xlevels # will print long list
dt_model$results
dt_model$finalModel # output is rpart
dt_model$bestTune
```

#### Plotting the decision tree
```{r fig.width=4, fig.height=4}
plot(dt_model)

##plot(dt_model$finalModel) # this will just plot an empty tree
plot(dt_model$finalModel, uniform=TRUE, branch=0.6, margin=0.1)
text(dt_model$finalModel, all=TRUE, use.n=TRUE)

library(rattle)
fancyRpartPlot(dt_model$finalModel)
```

#### Variable Importance
```{r fig.width=4, fig.height=4}
var_imp <- varImp(dt_model, scale=FALSE)
# by default varImp returns scaled results in the range 0-100, need to put scale=FALSE
print(var_imp)
# varImp(dt_model$finalModel) # output is a dataframe

plot(var_imp, top=10)
```

### Step 3 : Model Validation
#### Check the model performance with train_set

```{r}
#predict(dt_model)
#Error in eval(expr, envir, enclos) : object 'Sexmale' not found

pred_train1 <- predict(dt_model, newdata=train_set)
##pred_train1 <- predict(dt_model, newdata=train_set, type="raw") # same as above, default is "raw" for train object
#print(class(pred_train1)) # this is a factor variable
print(confusionMatrix(pred_train1, train_set$Survived))
## Accuracy : 0.8096

pred_train2_prob <- predict(dt_model$finalModel) # output is a matrix, it will output probability
#pred_train2_prob <- predict(dt_model$finalModel, type="prob") # same as above, default is prob for rpart object
print(head(pred_train2_prob))
# the deafult is to give probability, to get class of each observation, use type="class"
pred_train2_class <- predict(dt_model$finalModel, type="class")
print(confusionMatrix(pred_train2_class, train_set$Survived))
## Accuracy : 0.8096

# if you will apply to a newdata, gives problem
#predict(dt_model$finalModel, newdata=train_set)
#Error in eval(expr, envir, enclos) : object 'Sexmale' not found

# in caret train type is either raw or prob
table(valid_set$Survived)
pred_valid <- predict(dt_model, newdata=valid_set) # output is a factor variable
###pred_valid <- predict(dt_model, newdata=valid_set, type="raw") # same as above, default is raw for train object
table(pred_valid)
print(confusionMatrix(pred_valid, valid_set$Survived))
## Accuracy : 0.812

#Error in predict.train(dt_model, newdata = valid_set, type = "class") : 
#  type must be either "raw" or "prob"

```

```
# for train object, default value of type="raw" gives factor variable
# if you will give type="prob" will give problem:
# Error in .data.frame(out, , obsLevels, drop = FALSE) : 
#  undefined columns selected
# you will encounter this while calculating ROC curve
```

#### Check the model performance with valid_set

Do the missing value treatment, otherwise confusionMatrix will give this message.
```
Error in table(data, reference, dnn = dnn, ...) : 
  all arguments must have the same length
```

```{r}
table(valid_set$Survived)
pred_valid <- predict(dt_model, newdata=valid_set)
table(pred_valid)
confusionMatrix(pred_valid, valid_set$Survived)
## Accuracy : 0.812
```

### Step 4 : Calculate ROC Curve using ROCR package

In order to plot ROC curve, we need to calculate probabilities for each level of the response variable.
```
When you ask for class probabilities, model predictions are a data frame with separate columns for each class/level. If the response variable doesn't have levels that are valid column names, data.frame converts then to valid names. That creates a problem because the code is looking for a specific name but the data frame as a different (but valid) name.
```
0 and 1 are not valid R variable names, we get errors when generating class probabilities.
change them to "NO" and "Yes"
```
Error in `[.data.frame`(out, , obsLevels, drop = FALSE) : 
  undefined columns selected
```

```{r fig.width=4, fig.height=4}
levels(train_set$Survived)[levels(train_set$Survived)=="0"] <- "No"
levels(train_set$Survived)[levels(train_set$Survived)=="1"] <- "Yes"
levels(train_set$Survived)

levels(valid_set$Survived)[levels(valid_set$Survived)=="0"] <- "No"
levels(valid_set$Survived)[levels(valid_set$Survived)=="1"] <- "Yes"
levels(valid_set$Survived)

dt_model_prob <- train(Survived~., data=train_set, method="rpart")

library(ROCR)
# Calculate ROC Curve using ROCR package
pred_valid_prob <- predict(dt_model_prob, newdata=valid_set, type="prob") 
# this will output probabaility
roc_pred <- prediction(pred_valid_prob[, 2], valid_set$Survived)
roc_perf <- performance(roc_pred, measure="tpr", x.measure="fpr")

roc_auc <- performance(roc_pred, measure="auc", x.measure="cutoff")
auc <- unlist(roc_auc@y.values)
cat("*********** Value of AUC : ", auc, "\n") # AUC :  0.8107963

plot(roc_perf, colorize=T, main=paste("AUC: ", roc_auc@y.values), lwd=2)
abline(a=0, b=1, lwd=2)
```

### Step 5 : Apply the model on test dataset
```{r}
# dt_model$coefnames
#"Pclass"    "Sexmale"   "Age"       "SibSp"     "Parch"     "Fare"      "EmbarkedQ" "EmbarkedS"
# Cabin is not used in the model, so don't do anything.

colSums(is.na(test_final))
test_final$Age[is.na(test_final$Age)] <- median(test_final$Age, na.rm=TRUE)	
test_final$Fare[is.na(test_final$Fare)] <- median(test_final$Fare, na.rm=TRUE)	
colSums(is.na(test_final))
	
pred_test <- predict(dt_model, newdata=test_final)
dim(test_final) # 418 11
length(pred_test) # 331
###???? why the output differs from data rows??? may be due to NA values


table(pred_test)
submit_data <- data.frame(PassengerId = test_final$PassengerId, Survived = pred_test)
#write.csv(submit_data, file="output-DT-caret.csv", row.names=FALSE)

# make sure to check that there are no NA's
str(submit_data)
print(summary(submit_data))

#Error in data.frame(PassengerId = test_raw$PassengerId, Survived = pred_test) : 
#  arguments imply differing number of rows: 418, 331

```









